Cooperative Kernels: Safe Multitasking for Blocking Algorithms on GPUs

### 2016-11-10 write-ups/PLDI/paper.tex
Fxi latex compilation and move paragraph

\paragraph{OpenCL concepts}
An OpenCL program is typically executed by many \emph{threads} running
in parallel. Threads are organized into \emph{workgroups} of equal
size. The memory is organized in three levels: \emph{global} memory is
accessible by any thread, \emph{local} memory is shared at the workgroup
level (i.e., threads of different workgroups cannot access the same
local memory), and \emph{private} memory is dedicated to thread-level
storage and cannot be shared between threads. These concepts are
illustrated over the two examples.

\myfiglong\ref{fig:workstealing} illustrates an OpenCL implementation of
a work stealing application. All threads start their execution at the
function marked with the $\mathsf{kernel}$ keyword. They receive a
pointer to the task pools that reside in global memory, and which were
initialized by the host program beforehand to contain the initial
tasks. Each thread can identify the workgroup it belongs to via the
function \getgroupid, and OpenCL provides a series of primitives in the
same vein to access for instance the thread global id, or its local id
within a workgroup. Here the group id is stored into the variable
$\mathsf(pool\_id)$ (which is, by default, in private memory) and used
to access the relevant task pool. The $\mathsf(pop\_or\_steal)$ function
is called by all threads of a workgroup, although inside only the thread
with a local id equal to $0$, also called \emph{master thread} of the
workgroup, will try to pop or steal a task from the pools. The master
thread uses local memory inside $\mathsf(pop\_or\_steal)$ to ensure that
the returned value is valid for all threads of its workgroup. If a task
was obtained, then all threads of the workgroup can participate in the
processing of the task inside the $\mathsf{process\_task}$
function. \HEComment{Manage to say that a task is considered big enough
to benefit from parallel computation} If some new tasks are created
during the processing, it is again the master thread who will be in
charge of pushing them to the pool. Although not depicted here,
concurrent accesses to each pool is guarded by a mutex which is
implemented by atomic compare and swap operations on global
memory. Henceforth, master threads may be spinning on mutexes, and
workgroups are also spinning on the main loop, so a fair scheduling of
workgroups is desirable.
\begin{figure}

@@

Figure~\ref{fig:workstealing} shows a simplified version of an OpenCL
kernel that uses a work stealing queue.  \ADComment{Describe some
  aspects of the example, saying that the mutex is not a primitive,
  and can be coded up using the memory model but is not safe due to
  the execution model.  Make it clear that the lock function uses busy-waiting.}  Figure~\ref{fig:graphsearch} shows the
skeleton of a graph traversal algorithm.  \ADComment{Describe some
  aspects of the example, saying that the global barrier is not a
  primitive, and can be coded up using the memory model but is not
  safe due to the execution model.  Make it clear that the barrier uses busy-waiting.}

>>>>

% \paragraph{OpenCL concepts}
% An OpenCL program is typically executed by many \emph{threads} running
% in parallel. Threads are organized into \emph{workgroups} of equal
% size. The memory is organized in three levels: \emph{global} memory is
% accessible by any thread, \emph{local} memory is shared at the workgroup
% level (i.e., threads of different workgroups cannot access the same
% local memory), and \emph{private} memory is dedicated to thread-level
% storage and cannot be shared between threads. These concepts are
% illustrated over the two examples.

@@

Figure~\ref{fig:workstealing} shows a simplified version of an OpenCL
kernel that uses a work stealing queue.  \ADComment{Describe some
  aspects of the example, saying that the mutex is not a primitive,
  and can be coded up using the memory model but is not safe due to
  the execution model.  Make it clear that the lock function uses
  busy-waiting.}

\myfiglong\ref{fig:workstealing} illustrates an OpenCL implementation of
a work stealing application. All threads start their execution at the
function marked with the $\mathsf{kernel}$ keyword. They receive a
pointer to the task pools that reside in global memory, and which were
initialized by the host program beforehand to contain the initial
tasks. Each thread can identify the workgroup it belongs to via the
function $\getgroupid$, and OpenCL provides a series of primitives in the
same vein to access for instance the thread global id, or its local id
within a workgroup. Here the group id is stored into the variable
$\mathsf(pool\_id)$ (which is, by default, in private memory) and used
to access the relevant task pool. The $\mathsf(pop\_or\_steal)$ function
is called by all threads of a workgroup, although inside only the thread
with a local id equal to $0$, also called \emph{master thread} of the
workgroup, will try to pop or steal a task from the pools. The master
thread uses local memory inside $\mathsf(pop\_or\_steal)$ to ensure that
the returned value is valid for all threads of its workgroup. If a task
was obtained, then all threads of the workgroup can participate in the
processing of the task inside the $\mathsf{process\_task}$
function. \HEComment{Manage to say that a task is considered big enough
to benefit from parallel computation} If some new tasks are created
during the processing, it is again the master thread who will be in
charge of pushing them to the pool. Although not depicted here,
concurrent accesses to each pool is guarded by a mutex which is
implemented by atomic compare and swap operations on global
memory. Henceforth, master threads may be spinning on mutexes, and
workgroups are also spinning on the main loop, so a fair scheduling of
workgroups is desirable.

Figure~\ref{fig:graphsearch} shows the
skeleton of a graph traversal algorithm.  \ADComment{Describe some
  aspects of the example, saying that the global barrier is not a
  primitive, and can be coded up using the memory model but is not
  safe due to the execution model.  Make it clear that the barrier uses busy-waiting.}
